Verified Nanopasses for Compiling Conditionals

### 2025-06-20 paper.tex
finished section on explicate

The case for conditional expressions also requires a lemma about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove the following lemma which states that the ouput CFG of
\lstinline{explicate-tail} is an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}

@@

\noindent The case for conditional expressions needs a couple more
lemmas regarding control flow graphs. Similar to
\lstinline{explicate-tail}, we need a lemma which states that the
ouput CFG of \lstinline{explicate-assign} is an extension of its input
CFG.

\begin{lstlisting}
explicate-assign-CFG : ∀{x}{m}{t t'}{G₁ G₂}
  → explicate-assign x m t G₁ ≡ (t' , G₂)
  → G₁ ↝ G₂
\end{lstlisting}

@@

...

\begin{lstlisting}
explicate-pred-correct :
  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G$′$ G$″$ G$‴$}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
  $→$ ((s$′$ , $ρ′$) , G$′$ $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$ , $ρ″$))
  $→$ G$′$ $↝$ G$″$
  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G$″$ $≡$ (c$₁$ , G$‴$)
  $→$ (s , $ρ$) , G$‴$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}

>>>>

The case for conditional expressions also two lemmas about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove that the ouput CFG of \lstinline{explicate-tail} is
an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
\noindent And we prove that the big-step semantics is monotonic with
respect to the $↝$ relation on control flow graphs.
\begin{lstlisting}
eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
  → B$₁$ $↝$ B$₂$
  → (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  → (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
\end{lstlisting}

@@

\noindent The case for conditional expressions needs another lemma
regarding control flow graphs. Similar to \lstinline{explicate-tail},
we need a lemma which states that the ouput CFG of
\lstinline{explicate-assign} is an extension of its input CFG.

\begin{lstlisting}
explicate-assign-CFG : $∀${x}{m}{c c'}{G$₁$ G$₂$}
  → explicate-assign x m c G$₁$ $≡$ (c' , G$₂$)
  → G$₁$ $↝$ G$₂$
\end{lstlisting}

@@

The following is the statement of correctness for
\lstinline{explicate-pred-correct}.

\begin{lstlisting}
explicate-pred-correct :
  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G G$′$}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ if b then c$₂$ else c$₃$ $⇓$ v $⊣$ (s$″$,$ρ″$))
  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G $≡$ (c$₁$ , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}

\noindent This statement of correctness is interesting because it
requires the premise that either continuation \lstinline{c$₂$} or
\lstinline{c$₃$} evaluates to \lstinline{v}, depending on whether the
expression \lstinline{e$₁$} evaluates to \lstinline{true} or
\lstinline{false}. This explains the use of Agda's
\lstinline{if-then-else} in the second premise of
\lstinline{explicate-pred-correct}.\footnote{The formulation of this
lemma follows that of CompCert's \lstinline{transl\_condexpr\_prop}.}

The proof of \lstinline{explicate-pred-correct} uses many of the
monotonicity lemmas that we have already proved, plus one more.

\begin{lstlisting}
explicate-pred-CFG : $∀${e}{c$₁$ c$₂$ c}{G$₁$ G$₂$}
  $→$ explicate-pred e c$₁$ c$₂$ G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
